---
title: "Practical Machine Learning- Prediction Exercise"
output: html_document
---
Debo Okunrinboye

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data

The training data for this project are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

The test data are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]

The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har]. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## What you should submit

The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-). You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

## Approach:
Our outcome variable is classe, a factor variable. For this data set, “participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions: - exactly according to the specification (Class A) - throwing the elbows to the front (Class B) - lifting the dumbbell only halfway (Class C) - lowering the dumbbell only halfway (Class D) - throwing the hips to the front (Class E)

Two models will be tested using decision tree and random forest. The model with the highest accuracy will be chosen as our final model.

## Cross-validation

Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subtraining data (70% of the original Training data set) and subtesting (30%). Our models will be fitted on the subtraining data set, and tested on the subtesting data. Once the most accurate model is choosen, it will be tested on the original testing data set.

## Expected out-of-sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set. Our outcome variable “classe” is a factor variable. 

## Load required libraries

```{r}
library(lattice)
library(ggplot2)
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(e1071)
```

Set seed for reproducability
```{r}
set.seed(1234)
```

## Getting and Cleaning Data

Load the training and testing data set into  R and replace all missing values with "NA"
```{r}
training_data <- read.csv( "pml-training.csv", na.strings=c("NA","#DIV/0!", ""))
testing_data <- read.csv( "pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))
```

Delete columns with all missing values
```{r}
training_data <- training_data[,colSums(is.na(training_data)) == 0]
testing_data <- testing_data[,colSums(is.na(testing_data)) == 0]

```

Delete colums 1 to 7 to remove variables that are non-predictors in  training set
```{r}
training_data <- training_data[,-c(1:7)]
testing_data <- testing_data[,-c(1:7)]
dim(training_data)

```

Partition the training data set by assigning 70% into training and 30% into testing to allow for cross-validation.
```{r}
trainingpat <- createDataPartition(y=training_data$classe, p=0.70, list=FALSE)
subtraining <- training_data[trainingpat, ] 
subtesting <- training_data[-trainingpat, ]
```

# Decision Tree

Fit model on subtraining data.
```{r}
modelDT <- rpart(classe ~ ., data=subtraining, method="class")
predictionDT <- predict(modelDT, subtesting, type = "class")
rpart.plot(modelDT, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

Estimate the errors of the prediction algorithm in the Decision Tree model.
```{r}
confusionMatrix(predictionDT,as.factor(subtesting$classe))
```

# Random Forest

Fit model on subtraining data.
```{r}
modelRF <- randomForest(as.factor(classe) ~ ., data=subtraining, method="class")
predictRF <- predict(modelRF, subtesting, type = "class")
```

Estimate the errors of the prediction algorithm in the Random Forest model.
```{r}
RF_cm <- confusionMatrix(predictRF,as.factor(subtesting$classe))
RF_cm
plot(RF_cm$table, col = RF_cm$byClass, main = paste("Random Forest Accuracy Level =",round(RF_cm$overall['Accuracy'], 4)))
```

# Conclusion

## Result
The confusion matrices show that Random Forest algorithm performed better than Decision Trees. Random Forest model had an accuracy of 0.9956 or 99.65% (95% CI : (0.9935, 0.9971)) as compared to Decision Tree model with  accuracy of 0.7541 or 75.41% (95% CI : (0.7429, 0.7651)). The Random Forests model is the model of choice. The expected out-of-sample error is estimated at 0.005, or 0.5%.


## Expected out-of-sample error
The expected out-of-sample error is estimated at 0.005, or 0.5% calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

## Submission
Predict outcome levels on the original testing data set using Random Forest algorithm
```{r}
predictfinal <- predict(modelRF, testing_data, type="class")
predictfinal
```








